Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-21 _tex/index.tex
Built site for gh-pages

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. This research primarily focuses on three: the over-reliance
on Thurstone's Case V assumptions in the statistical analysis of CJ
data, the apparent disconnect between CJ's trait measurement and
hypothesis testing, and the unclear role of comparison algorithms on the
method's reliability and validity. The following sections will discuss
each of these issues in detail, followed by the introduction of a
theoretical model and its statistical translation, which aims to address
all three concerns simultaneously.

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects the variability in the stimulus' perceived trait
level. Figure~\ref{fig-discriminal_process} shows example distributions
of discriminal process for two stimuli (objects).

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}

@@

\caption{\label{fig-discriminal_process}Example distributions of
discriminal processes for two stimuli (objects). Extracted from
\citet[pp.~249]{Bramley_2008}.}

@@

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable except through comparison, the \emph{law of
comparative judgment} becomes essential. This law asserts that when
assessing a specific trait by comparing two stimuli, the stimulus
positioned further along the continuum is perceived as having a higher
level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the stimuli's
discriminal processes, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows an example distribution of
the discriminal difference for two stimuli (objects).

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}

\caption{\label{fig-comparative_judgment}Distribution of the discriminal
difference between two stimuli (objects). Extracted from
\citet[pp.~251]{Bramley_2008}.}

Importantly, the theory's general form primarily addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
applicability, Thurstone introduced five distinct cases, each defined by
progressively simplifying assumptions. Table~\ref{tbl-thurstone_cases}
summarizes these cases, focusing on key assumptions such as the
distribution of discriminal processes, the similarity of discriminal
dispersions across stimuli, the correlation between stimuli, and which
judges perform the comparisons. For a comprehensive discussion of this
progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248-253]{Bramley_2008}.

@@

\subsection{The Case V and the statistical analysis of CJ
data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}

Surprisingly, despite its reliance on the largest number of simplifying
assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors Case V's
assumptions, with one key difference: while Case V assumes a Normal
distribution for the stimuli's discriminal processes, the BTL model uses
the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little effect on
the model's estimation or interpretation, as the Normal and Logistic
distributions differ by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
Figure~\ref{fig-logistic_vs_normal}).

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}

@@

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
As a result, its assumptions may not suit applications beyond the
psycho-physical contexts for which it was created. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
assumptions could prove problematic with more complex, less homogeneous
stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
applications often deal with complex, less homogeneous stimuli, two key
assumptions of Case V may not consistently hold in theory or practice:
the zero correlation and equal dispersion between stimuli.

The assumption of \emph{zero correlation between stimuli} is best
illustrated with an example. For instance, when evaluating text quality,
this assumption suggests that a judge's perception of quality in one
text does not influence their perception of the same trait in the
comparison text. Thurstone attributes this independence to the
cancellation of potential judges' biases, driven by opposing and equally
weighted factors; commonly known as `mood' and `simultaneous contrast'
effects, that operate during pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
mathematically demonstrated by \citet{Andrich_1978}, assuming additive
biases and a logit scale derived from the BTL model.

However, given that evidence has shown judges' quality assessments are
influenced by multiple, complex aspects of the texts
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
plausible that the accuracy of these judgments also varies based on
factors such as age, culture, education, expertise
\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
individual judge themselves
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.

{This means that the cancellation of judges' biases may not hold in
theory or practice, and that judges' biases may depend on the stimuli,
by means of an interaction (but present) or because the judges works as
a confounder}

Such differences may not be detected through analyses of bias and misfit
\citep[pp.~683]{Kelly_et_al_2022}.

\subsection{The disconnect between trait measurement and hypothesis
testing}\label{the-disconnect-between-trait-measurement-and-hypothesis-testing}

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses or hypotheses tests. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can introduce bias into the
analysis and reduce the precision of hypothesis tests. Notably, the
direction and magnitude of the bias are often unpredictable; results may
be attenuated, exaggerated, or remain unaffected, depending on the
amount of uncertainty present in the scores and the actual effects being
tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, reduced precision in hypothesis tests weakens their
statistical power, ultimately increasing the likelihood of committing
type-I or type-II errors \citep{McElreath_2020}.

@@

\subsection{The role and impact of comparison
algorithms}\label{the-role-and-impact-of-comparison-algorithms}

\section{Theory}\label{sec-theory}

\subsection{A theoretical model for CJ}\label{sec-theory-theoretical}

@@

\textbf{Non-financial interests:} Author XX serve on advisory broad of
Company Y but receives no compensation this role.

@@

\textbf{CRediT authorship contribution statement:}
\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
J.M.R.E.; \emph{Writing - review \& editing:} S.G., S.DM., and T.vD.;
\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
\emph{Project administration:} S.G. and S.DM.; \emph{Funding
acquisition:} S.G. and S.DM.

>>>>

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity suggests that scores generated by CJ can
accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023},
while research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. The present study primarily focuses on three: the
over-reliance on the assumptions of Thurstone's Case V in the
statistical analysis of CJ data, the apparent disconnect between CJ's
trait measurement and hypothesis testing, and the unclear role of the
diverse assessment design features on CJ's reliability and validity. The
following sections begin with a brief overview of Thurstone's theory and
a detailed discussion of these issues. Subsequently, the study
introduces a theoretical model for CJ that builds upon Thurstone's
theory, alongside its statistical translation, designed to address all
three concerns simultaneously.

\section{Thurstone's theory}\label{sec-thurstone_theory}

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} suggests that two factors determine the
dichotomous outcome of pairwise comparisons: the discriminal process of
each stimulus and their discriminal difference. The \emph{discriminal
process} refers to the psychological effect each stimulus exerts on the
judges, or more simply, the underlying perception of the stimulus' trait
level. According to the theory, the discriminal process for each
stimulus follows a Normal distribution. The mode (mean) of this
distribution, referred to as the \emph{modal discriminal process},
represents the stimulus' position on the trait continuum. Meanwhile, the
dispersion of the distribution, referred to as the \emph{discriminal
dispersion}, reflects the variability in the perceived trait level of
the stimulus.

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This suggests that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts can be illustrated through an example. For instance, in
the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would favor text A.

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}

@@

\subcaption{\label{fig-discriminal_process}Discriminal processes}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}

\subcaption{\label{fig-discriminal_difference}Discriminal difference}

\caption{\label{fig-thurstone_theory}Example distribution of discriminal
processes and their discriminal difference for two written texts
(stimuli or objects). Extracted from \citet[pp.~249-251]{Bramley_2008}.}

@@

Importantly, the general form of Thurstone's theory primarily addressed
pairwise comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Thus, for practical application,
Thurstone introduced five distinct cases derived from this general form,
each defined by progressively simplifying assumptions.
Table~\ref{tbl-thurstone_cases} summarizes these cases, highlighting key
assumptions such as the distribution of discriminal processes, the
similarity of discriminal dispersions across stimuli, the correlation
between stimuli, and the number of judges performing the comparisons.
For a comprehensive discussion of this progression, refer to
\citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

\subsection{The Case V and the statistical analysis of CJ
data}\label{sec-theory-issue1}

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of stimuli, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions differ by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}

@@

\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}

@@

However, Case V was originally developed to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve these types of traits and stimuli, two
main assumptions of Case V may not consistently hold in theory or
practice: the zero correlation and equal dispersion between stimuli.

The assumption of \emph{zero correlation between stimuli} is more easily
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption suggests that a
judge's perception of the trait in one text does not influence his
perception of the same trait on another. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, assuming
discriminal processes with additive biases. However, it is easy to
imagine two scenarios where the zero correlation assumption almost
certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli, and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, evidence suggests that when judges are confronted
with multi-dimensional, complex traits and heterogeneous stimuli, such
as when evaluating written text quality, their assessments are
influenced by multiple, intricate aspects of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
such cases, it is not inconceivable that these various aspects of the
stimuli unevenly influence the judges' perception of one text over
another, leading to biases that resist cancellation, as the effects of
the factors involved in the comparison are neither equally weighted nor
opposing. For example, this may occur when a judge, while assessing the
argumentative quality of a text, places disproportionate emphasis on
grammatical errors, ultimately favoring texts with fewer errors but
weaker arguments. Although the literature does not provide direct
evidence for this specific argument, studies such as
\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
supporting the idea that the factors involved in pairwise comparisons
may not always cancel out.

\subsection{The disconnect between trait measurement and hypothesis
testing}\label{sec-theory-issue2}

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses and tests, or to make
decisions regarding the exclusion of certain data in these analyses and
tests. The literature shows that these scores have been employed to
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023} or to test hypotheses
related to the underlying traits of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
Additionally, the BTL scores have been used to detect biases in judges'
ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, as well as to
identify ``misfit'' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, with
considerations for their possible exclusion.

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses and tests while warning against
eliminating data through ad-hoc univariate procedures. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can bias the analysis and
reduce the precision of hypothesis tests. Notably, the direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Moreover, eliminating data using ad-hoc univariate procedures using
these scores compounds this issue by excluding possible valuable
information from the analyses \citep{McElreath_2020}, further
introducing bias. Finally, the reduced precision in hypothesis tests
weakens their statistical power, increasing the likelihood of type-I or
type-II errors \citep{McElreath_2020}.

@@

\subsection{Assessment design features and their role on reliability and
validity}\label{sec-theory-issue3}

\section{An updated theoretical and statistical model for
CJ}\label{sec-theory}

\subsection{The theoretical model}\label{sec-theory-theoretical}

@@

\textbf{Non-financial interests:} The authors have no relevant
non-financial interest to disclose.

@@

\textbf{CRediT authorship contribution statement:}
\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
J.M.R.E.; \emph{Writing - review and editing:} S.G., S.DM., and T.vD.;
\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
\emph{Project administration:} S.G. and S.DM.; \emph{Funding
acquisition:} S.G. and S.DM.
